A primer to numerical simulations: The perihelion motion of Mercury

### 2017-07-01 tex/MercuryNumerical.tex
Worked through section 1-3

@@

Numerical simulations play a key role in modern physics for they allow one to tackle theoretical problems
not accessible otherwise, e.g., since there are too many particle participating in the system (as in
simulations for weather predictions) or the interactions are too complicated to allow for a systematic,
perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level). 
 The goal of this paper is it to
introduce a project that could be used to make numerical simulations themselves the topic of the class.
On the example of the perihelion motion of the planet Mercury the students are supposed to get in touch with
\begin{itemize}
\item the importance of differential equations in theoretical physics;
\item the numerical implementation of Newtonian dynamics;
\item systematic tests and optimization of computer codes;
\item effective tools to estimate the result a priori as an important cross check;
\item the visualization of numerical results using V-Phython.
\end{itemize}
The course as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the 
concepts of differential equations their discretization is discussed based on Newtons law of gravitation and possible
extensions thereof. Afterwards the visualization of the resulting trajectories using V-Phython is introduced and
applied to the problem at hand. In particular tools are developed to extract the relevant quantity from the result
of the simulation.
Finally the principle of dimensional analysis is presented as a tool to cross check, if the results of the simulation
are sensible.

@@

We can say that we understand a physical system, if we can demonstrate that the assumed force leads to the trajectories
observed. In other words, we need to show that we can calculate the location in space of the object of interest at any point
in time, once the initial conditions are fixed properly. If we can neglect the finite size of this object (and
in particular its orientation in space) the location is parametrized
by a single three vector $\vec x(t)$. As will become clear below in order to describe and control the dynamics of a physical
object in addition we need to be able to calculate its velocity, $\vec v(t)$, related to the location via
\begin{equation}\label{eq:x_update}
\vec x(t+\Delta t) = \vec x(t) + \vec v(t) \Delta t + ... \ , \label{eq:vdef}
\end{equation}
for some infinitesimally small $\Delta t$ and the acceleration, $\vec a(t)$ that describes the change of the velocity
\begin{eqnarray}\label{eq:v_update}
\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + ...\ . \label{eq:adef}
\end{eqnarray}
The dots in the above expressions indicate that there are in general additional terms that may be expressed with
higher powers in $\Delta t$, however, for sufficiently small $\Delta t$ those can be safely neglected.
Thus we may define the time derivative via
\begin{equation}
\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec x(t)}{\Delta t} =: \frac{d\vec x(t)}{dt}  = \dot{\vec  x}(t) \ ,
\end{equation}
where $\Delta \vec x(t)=\vec x(t+\Delta t)-\vec x(t)$ and we introduced with the last expression a common short hand notation
for time derivatives. Analogously we get
\begin{eqnarray}
\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t) \ , \\
 &=& \frac{d^2\vec x(t)}{dt^2}  = \ddot{\vec x}(t) \ ,
\end{eqnarray}
where we introduced in the second line the second derivative.

It was Newton who observed that if a body is at rest it will remain at rest, and if it is in motion it will remain in motion in a 
constant velocity in a straight line, unless it is acted upon by some force --- this is known as Newtons first law. 
Said differently: a force $\vec F$ shows up by changing the motion of some object. This is quantified in Newton's second law
\begin{equation}
\vec F(\vec x, t) = \frac{d}{dt}(m \vec v) \ .
\end{equation} 
If the mass does not change~\footnote{A well known example where $m$ does change with time is a 
rocket, whose mass decreases as the rocket rises.} with time this reduces to the well known
\begin{equation}
\vec F(\vec x) = m \vec a(t) = m\dot{\vec v}(t) = m\ddot{\vec{x}}(t) \ . \label{eq:newton2}
\end{equation} 
Note that in general the force could depend also on the time or the velocity. We here restrict ourselves to the
case relevant for our example where the force depends on the location only. 
Therefore, 
as soon as the force $\vec F(\vec x)$ is known for all $\vec x$ one can in principle calculate the trajectory, e.g. solving eq.~(\ref{eq:newton2})
for $\vec x(t)$.
Sometimes this requires some advanced knowledge in math, sometimes no closed form solution exists. However, alternatively 
one can calculate the whole trajectory
of some test body that experiences this force by a successive application of the rules 
given in Eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
\begin{enumerate}
\item For a given time $t$, where $\vec x(t)$ and $\vec v(t)$ are known, use Eq.~(\ref{eq:newton2}) to calculate $\vec a(t)$.
\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec x(t+\Delta t)$ and
\item then use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
\item Go back to (i) with $t\to t+\Delta t$.
\end{enumerate}
Clearly to initiate the procedure at some time $t_0$ both $\vec x(t_0)$ as well as $\vec v(t_0)$ must be known --- the 
trajectories depend on these initial conditions.~\footnote{In general a differential equation of $n^{\rm th}$ degree (where
the highest derivative is $n$) needs $n$ initial conditions specified. For $n=2$ those are often chosen as location and velocity
at some defined time, but also to pick two location at different times is possible. }

Clearly for this procedure to work $\Delta t$ must be sufficiently small. What this means depends on the process
studied. One way to estimate, if $\Delta t$ is small enough, is to verify, if the relation
\begin{equation}
|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} F(\vec x(t))\Delta t\ . 
\label{eq:check}
\end{equation}
holds, since the next term neglected in Eq.~(\ref{eq:vdef}) reads $(1/2)a(t)(\Delta t)^2$.  Eq.~({\ref{eq:check}})
also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since
the time steps need to be small enough that all changes induced by the force get resolved. Clearly,
a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of
the solutions: A valid result has to be insensitive the concrete value $\Delta t$ chosen --- in particular replacing
$\Delta t$ by $\Delta t/2$ should not change the result significantly.

@@

For this concrete example the starting point for the force is Newtons law
of gravitation
\begin{equation}
\vec F_N(\vec x) = \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s${-2}$ is the Newtonian constant of gravitation,
$m$ is the mass of Mercury and $M_\odot=2\times 10^{30}$ kg is the mass of the sun.
In addition $r=|\vec x(t)|$ denotes the distance between sun and Mercury, when we assume that the sun
is infinitely more heavy than the planet and located at the center of the coordinate system. 
Although this is not exact, since $m/M_\odot\sim 10^{-8}$ this is
a pretty good approximation. For later convenience we introduce the Schwarzschildradius of the sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 3 \ \mbox{km} \ ,
\end{equation}
where $c=3\times 10^8$ m/s denotes the speed of light.
Note the $r_S$ is the characteristic length scale of the gravitational field of the sun
for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and
$c$. This is what we need $r_S$ for later in the discussion. That the Schwarzschild radius is
also an important quantity to characterize black holes is not relevant for the discussion
at hand.
With this Newtons second law reads 
\begin{equation}
\ddot{\vec x} = \frac{c^2}{2}\left(\frac{r_S}{r^2}\right) \ .
\label{eq:newton}
\end{equation}
%\begin{itemize}
%\item{\bf First task:}
%Now the students should implement the program outlined in Sec.~\ref{sec:tva} and visualize the resulting
%trajectories of Mercury for different start parameters.
%Why is it sufficient to work in two dimensions using simply 
%$\vec x(t)=(x(t),y(t))^T$?
% When the parameters are chosen appropriately close
%orbits should emerge that are fixed in time --- in particular the perihelion (the point of closest approach
%of planet and sun) stays fixed in space.
%\end{itemize}
In general an attractive force that vanishes at large distances leads,
 depending on the initial conditions,
  either to closed orbits or open orbits --- 
  in the latter case the planet simply disappears from the sun. Depending on time available
for the course  it might be helpful to explain to the students the origin of this, however, since the most
transparent reasoning needs an integration as well as the concept of angular momentum
not necessarily familiar to the whole class (details are given in the appendix {\bf to be done}), 
it appears in general more appropriate to work out together with the students 
that a moving body can not be kept by a force if its momentum is too large.
 Then they may try to find closed orbits in the simulation, e.g., by varying
the start velocity while keeping the start location fixed.

The closed orbits that emerge from a force that scales as $1/r^2$
are elliptic and in particular fixed in space. In particular the point of closest approach of the planet
to the sun, the perihelion, does not move.
  However, as soon
as the potential is different, the perihelion moves. Because of this the behavior of the perihelion
is a very sensitive probe of the gravitational potential.

The observed perihelion motion of Mercury is nowadays determined as $(574.10\pm 0.65)''$ per
100 earth years~\footnote{The symbol $''$ denotes 'arc seconds': 1$''=(1/3600)^o$. }, but it should be noted that the bulk of this number
can be understood by the presence of other planets within the Newtonian theory, since
their gravitational force also acts on Mercury. However, a
residual motion of $(43)''$ in 100 earth years remained unexplained, until Einstein quantified the
predictions of General Relativity to this particular observable.

{\bf Here should be a brief description of the relation between Newtons dynamics and General Relativity}

To allow for a movement of the perihelion also in the study presented here the
potential of Eq.~(\ref{eq:newton}). We do this using the following ansatz
\begin{equation}
\ddot{\vec x} = \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\left(\frac{r_S}{r}\right)\right) \ ,
\label{eq:newton_art}
\end{equation}
where $\alpha$ is some parameter. The idea is that the correction must be dimension less and
should scale with some inverse power of $r$, for we still need to demand that the potential 
vanishes at large distances. Since the only parameter of the system with dimension
of a length is $r_S$, the correction should scale as $(r_S/r)$. When replacing $r$ by a typical
distance of Mercury to the sun, $(r_S/r)\sim 10^{-7}$, it should be sufficient to include this kind
of term to first order only. For comparison: the corresponding number for the venus, the neighbor
of  Mercury, reads ...., which naturally explains why the effect is observable for Mercury only (IS IT - CHECK!)
%\end{itemize}

\section{Numerical Implementation}
\label{sec:Numerical Implementation}
In this section we present the numerical implementation as well as the visualization of planetary trajectories and in
particular
  the perihelion motion of Mercury.  We select \python~\cite{} as  programming language, because \python is easy to learn, 
  intuitive to understand and an open source language\footnote{See also appendix \ref{appendix:python} for an instruction 
  how to set up python on different operating systems.}.  No prior knowledge of \python or any other programming language 
  is required because we explain all the steps necessary to create the simulation.  We also provide a working 
  example \cite{} that can be used as template.

@@

In python, this reads
\begin{lstlisting}
	# Definition of parameters
	rM0 = 4.6    # in units of R0
	vM0 = 0.34   # in units of R0/T0
	aM  = 0.99   # in units of R0/T0**2
\end{lstlisting}
So far we only fixed the length of the vectors.
Next, we set up the initial directions, which will describe the motion in the two-dimensional plane.  We will build on the existing \python module \texttt{Vpython}, which provides  an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.  Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of mercury is perpendicular to the vector which connects mercury and the sun (see figure \ref{fig:sun_merc}):
\begin{lstlisting}
	# Import the class vector from python
	from vpython import vector
	# Initialize distance and velocity vectors
	vec_rM0 = vector(0, rM0, 0)
	vec_vM0 = vector(vm0, 0, 0)
\end{lstlisting}
According to Eq.~(\ref{eq:newton}), the force which acts on mercury changes the velocity of mercury which eventually changes the position. However, we first need to fix the time step $\Delta t$ and use the estimate
of Eq.~(\ref{eq:check}) as a guidance {\bf Careful: Check units of dt}
\begin{lstlisting}
	# Definition of the time step
        dt = 2 * vm0 / aM / 100
\end{lstlisting}
Here the factor $1/100$ makes sure that $\Delta t$ is indeed
consistent with Eq.~(\ref{eq:check}).  {\bf We need to make sure that 'dt' is known
to the subroutines}
Now we are in the position to calculate
location and velocity of the planet at $t_0+\Delta t$ using the following comands
\begin{lstlisting}
	# Compute the strength of the acceleration
	aMS = aM * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
	# Multiply by the direction
	vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
	# Update velocity vector
	vec_vM_new = vec_vM_old + vec_aMS * dt
	# Update position vector
	vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note the beauty of working with the predefined \texttt{vector} class: the basic vector operations are already implemented.  The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.  Also the magnitude of a vector -- \texttt{vector.mag} -- is a property of the vector and can be easily extracted.

@@

Suppose we want to save the number of turns in the variable $counter$, the maximal number of turns in $maxcounter$ (here e.g. 10), the initial aphelion in the vector $a0$ and the final aphelion in the vector $a$. In addition to find out when the aphelion is reached, the last two positions $rlast$ and $rbeforelast$ have to be known. To set this up, before the while loop the following code needs to be included
\begin{lstlisting}
	r=r0
	rlast=r0
	counter=-1
	maxcounter=10
\end{lstlisting}
Then the following code needs to be included in the $while$-loop
\begin{lstlisting}
    rbeforelast=rlast
    rlast=r
    r=r+v*dt
    if (counter==-1): a0=rlast
    if (rbeforelast.mag<rlast.mag and rlast.mag>r.mag): counter=counter+1
    if counter==maxcounter:
    	a=rlast
    	break
\end{lstlisting}
and the condition for the $while$-loop should be set to $True$.
The angle between two vectors is given by the formula
 \begin{equation}
 	\sphericalangle(\vec{a}_0,\vec{a}) = \cos^{-1} \left( \frac{\vec{a}_0 \cdot \vec{a}}{|\vec{a}_0|\:|\vec{a}|} \right)
 \end{equation}
This is readily implemented in Vpython via
\begin{lstlisting}
	angle=acos(dot(a0,a)/(a0.mag*a.mag))/maxcounter*180/pi
	print("angle: ", (angle))
\end{lstlisting}

@@

Instead of providing the template to the students, they could build the program from scratch. Of course, this requires some basic knowledge in VPython, as they could acquire for example by working through a ball in the box example or the like. They could even look up the needed parameters on their own.
\item \textbf{Why can we work in one plane?}

@@

\appendix
\section{Instructions to install V-phython}
\label{appendix:python}

>>>>

\usepackage[prependcaption, textsize=tiny]{todonotes}

@@

Numerical simulations play a key role in modern physics for they allow one to tackle theoretical problems
not accessible otherwise, e.g., since there are too many particle participating in the system (as in
simulations for weather predictions) or the interactions are too complicated to allow for a systematic,
perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level). 
 The goal of this paper is it to
introduce a project that could be used to make numerical simulations themselves the topic of the class.
On the example of the perihelion motion of the planet Mercury the students are supposed to get in touch with
\begin{itemize}
\item the importance of differential equations in theoretical physics;
\item the numerical implementation of Newtonian dynamics;
\item systematic tests and optimization of computer codes;
\item effective tools to estimate the result a priori as an important cross check;
\item the visualization of numerical results using \texttt{VPython}.
\end{itemize}
The course as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the 
concepts of differential equations their discretization is discussed based on Newtons law of gravitation and possible
extensions thereof. Afterwards the visualization of the resulting trajectories using \texttt{VPython} is introduced and
applied to the problem at hand. In particular tools are developed to extract the relevant quantity from the result
of the simulation.
Finally the principle of dimensional analysis is presented as a tool to cross check, if the results of the simulation
are sensible.

@@

We can say that we understand a physical system, if we can demonstrate that the assumed force leads to the trajectories
observed. In other words, we need to show that we can calculate the location in space of the object of interest at any point
in time, once the initial conditions are fixed properly. If we can neglect the finite size of this object (and
in particular its orientation in space) the location is parametrized
by a single three vector $\vec x(t)$. As will become clear below, in order to describe and control the dynamics of a physical
object in addition we need to be able to calculate its velocity, $\vec v(t)$, related to the location via
%
\todo{I see here a potential conflict with the implementation because we use $x(t-\Delta t)$ in order to get $v$ and so on}
%
\begin{equation}\label{eq:x_update}
\vec x(t+\Delta t) = \vec x(t) + \vec v(t) \Delta t + ... \ , \label{eq:vdef}
\end{equation}
for some infinitesimally small $\Delta t$ and the acceleration, $\vec a(t)$ that describes the change of the velocity
\begin{eqnarray}\label{eq:v_update}
\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + ...\ . \label{eq:adef}
\end{eqnarray}
The dots in the above expressions indicate that there are in general additional terms that may be expressed with
higher powers in $\Delta t$, however, for sufficiently small $\Delta t$ those can be safely neglected.
Thus we may define the time derivative via
\begin{equation}
\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec x(t)}{\Delta t} =: \frac{d\vec x(t)}{dt}  = \dot{\vec  x}(t) \ ,
\end{equation}
where $\Delta \vec x(t)=\vec x(t+\Delta t)-\vec x(t)$ and we introduced with the last expression a common short hand notation
for time derivatives. Analogously we get
\begin{eqnarray}
\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t) \ , \\
 &=& \frac{d^2\vec x(t)}{dt^2}  = \ddot{\vec x}(t) \ ,
\end{eqnarray}
where we introduced in the second line the second derivative.

It was Newton who observed that if a body is at rest it will remain at rest, and if it is in motion it will remain in motion in a 
constant velocity in a straight line, unless it is acted upon by some force --- this is known as Newtons first law. 
Said differently: a force $\vec F$ shows up by changing the motion of some object. This is quantified in Newton's second law
\begin{equation}
\vec F(\vec x, t) = \frac{d}{dt}(m \vec v) \ .
\end{equation} 
If the mass does not change~\footnote{A well known example where $m$ does change with time is a 
rocket, whose mass decreases as the rocket rises.} with time this, reduces to the well known
\begin{equation}
\vec F(\vec x) = m \vec a(t) = m\dot{\vec v}(t) = m\ddot{\vec{x}}(t) \ . \label{eq:newton2}
\end{equation} 
Note that in general the force could depend also on the time or the velocity. We here restrict ourselves to the
case relevant for our example where the force depends on the location only. 
Therefore, 
as soon as the force $\vec F(\vec x)$ is known for all $\vec x$ one can in principle calculate the trajectory, e.g. solving eq.~(\ref{eq:newton2})
for $\vec x(t)$.
Sometimes this requires some advanced knowledge in math, sometimes no closed form solution exists. However, alternatively 
one can calculate the whole trajectory
of some test body that experiences this force by a successive application of the rules 
given in Eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
%
\todo{See previous comment.}
%
\begin{enumerate}
\item For a given time $t$, where $\vec x(t)$ and $\vec v(t)$ are known, use Eq.~(\ref{eq:newton2}) to calculate $\vec a(t)$.
\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec x(t+\Delta t)$ and
\item then use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
\item Go back to (i) with $t\to t+\Delta t$.
\end{enumerate}
Clearly to initiate the procedure at some time $t_0$ both $\vec x(t_0)$ as well as $\vec v(t_0)$ must be known --- the 
trajectories depend on these initial conditions.~\footnote{In general a differential equation of $n^{\rm th}$ degree (where
the highest derivative is $n$) needs $n$ initial conditions specified. For $n=2$ those are often chosen as location and velocity
at some defined time, but also to pick two location at different times is possible. }

Clearly for this procedure to work $\Delta t$ must be sufficiently small. What this means depends on the process
studied. One way to estimate, if $\Delta t$ is small enough, is to verify, if the relation
\begin{equation}
|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} |\vec F(\vec x(t)) |\Delta t\ . 
\label{eq:check}
\end{equation}
holds, since the next term neglected in Eq.~(\ref{eq:vdef}) reads $(1/2)a(t)(\Delta t)^2$.  Eq.~({\ref{eq:check}})
also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since
the time steps need to be small enough that all changes induced by the force get resolved. Clearly,
a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of
the solutions: A valid result has to be insensitive the concrete value $\Delta t$ chosen --- in particular replacing
$\Delta t$ by $\Delta t/2$ should not change the result significantly.

@@

For this concrete example the starting point for the force is Newtons law
of gravitation
\begin{equation}
\vec F_N(\vec x) = \frac{G_N m M_\odot}{r^2} \frac{\vec x}{r}\ ,
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s${-2}$ is the Newtonian constant of gravitation,
$m$ is the mass of Mercury and $M_\odot=2\times 10^{30}$ kg is the mass of the sun.
In addition $r=|\vec x(t)|$ denotes the distance between sun and Mercury, when we assume that the sun
is infinitely more heavy than the planet and located at the center of the coordinate system. 
Although this is not exact, since $m/M_\odot\sim 10^{-8}$ this is a pretty good approximation.
%
\todo{This possibly can become a problem for our accuracy because it is at the 10\% level of our expected result, right? Same for $c$.}
%
For later convenience we introduce the Schwarzschildradius of the sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 3 \ \mbox{km} \ ,
\end{equation}
where $c=3\times 10^8$ m/s denotes the speed of light.
Note the $r_S$ is the characteristic length scale of the gravitational field of the sun
for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and
$c$. This is what we need $r_S$ for later in the discussion. That the Schwarzschild radius is
also an important quantity to characterize black holes is not relevant for the discussion
at hand.
With this Newtons second law reads 
\begin{equation}
\ddot{\vec x} = \frac{c^2}{2}\left(\frac{r_S}{r^2}\right) \ .
\label{eq:newton}
\end{equation}
%\begin{itemize}
%\item{\bf First task:}
%Now the students should implement the program outlined in Sec.~\ref{sec:tva} and visualize the resulting
%trajectories of Mercury for different start parameters.
%Why is it sufficient to work in two dimensions using simply 
%$\vec x(t)=(x(t),y(t))^T$?
% When the parameters are chosen appropriately close
%orbits should emerge that are fixed in time --- in particular the perihelion (the point of closest approach
%of planet and sun) stays fixed in space.
%\end{itemize}
In general an attractive force that vanishes at large distances leads,
%
\todo{I am confused. I thought only $1/r$ and the harmonic oscillator potential lead to closed orbits (Bertrand's theorem).}
%
 depending on the initial conditions, either to closed orbits or open orbits --- in the latter case the planet simply disappears from the sun.
Depending on time available for the course  it might be helpful to explain to the students the origin of this, however, since the most transparent reasoning needs an integration as well as the concept of angular momentum not necessarily familiar to the whole class (details are given in the appendix 
%
\todo{\bf Include this appendix part...}
%
), 
it appears in general more appropriate to work out together with the students that a moving body can not be kept by a force if its momentum is too large.
Then they may try to find closed orbits in the simulation, e.g., by varying the start velocity while keeping the start location fixed.

The closed orbits that emerge from a potential that scales as $1/r$ are elliptic and fixed in space.
In particular the point of closest approach of the planet to the sun, the perihelion, does not move.
However, as soon as the potential deviates from $1/r$, the perihelion moves. Because of this, the behavior of the perihelion
is a very sensitive probe of the gravitational potential.

The observed perihelion motion of Mercury is nowadays determined as $(574.10\pm 0.65)''$ per
100 earth years~\footnote{The symbol $''$ denotes 'arc seconds': 1$''=(1/3600)^o$. }, but it should be noted that the bulk of this number
can be understood by the presence of other planets within the Newtonian theory, since
their gravitational force also acts on Mercury. However, a
residual motion of $\delta \phi_M = (43)''$ in 100 earth years remained unexplained, until Einstein quantified the
predictions of General Relativity to this particular observable.

%
\todo{\bf Here should be a brief description of the relation between Newtons dynamics and General Relativity}
%

We allow for a movement of the perihelion by modifying the potential of Eq.~(\ref{eq:newton}) with the following ansatz
\begin{equation}
\ddot{\vec x} = \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\frac{r_S}{r}\right) \ ,
\label{eq:newton_art}
\end{equation}
where $\alpha$ is some parameter related to the effects by general relativity.
The idea is that the correction must be dimension-less and should scale with some inverse power of $r$, for we still need to demand that the potential vanishes at large distances.
Since the only parameter of the system with dimension of a length is $r_S$, the correction should scale as $(r_S/r)$.
When replacing $r$ by a typical distance of Mercury to the sun, $(r_S/r)\sim 10^{-7}$, it should be sufficient to include this kind of term to first order only.
For comparison: the corresponding number for the Venus, the neighbor of Mercury, reads $r \sim 5 \cdot 10^{-8}$.
Therefore the expected ratio of residual motion, favoring in the orbit period $T$, would result in
\begin{equation}
	\delta \phi_V \approx \frac{1}{T_V} \delta \phi_M T_M \frac{r_M}{r_V} \approx 9'' \, ,
\end{equation}
which precisely describes the experimental value of $9''$.

%\end{itemize}

\section{Numerical Implementation}
\label{sec:Numerical Implementation}
In this section we present the numerical implementation as well as the visualization of planetary trajectories and in
particular
  the perihelion motion of Mercury.  We select \python~\cite{} as  programming language, because \python is easy to learn, 
  intuitive to understand and an open source language\footnote{See also appendix \ref{appendix:python} for an instruction 
  how to set up \python on different operating systems.}.  No prior knowledge of \python or any other programming language 
  is required because we explain all the steps necessary to create the simulation.  We also provide a working 
  example \cite{} that can be used as template.

@@

In \python, this reads
\begin{lstlisting}
	# Definition of parameters
	rM0 = 4.6    # in units of R0
	vM0 = 0.34   # in units of R0/T0
	aM  = 0.99   # in units of R0/T0**2
\end{lstlisting}
So far we only fixed the length of the vectors.
Next, we set up the initial directions, which will describe the motion in the two-dimensional plane.  We will build on the existing \python module \texttt{Vpython}, which provides  an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.  Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of mercury is perpendicular to the vector which connects mercury and the sun (see figure \ref{fig:sun_merc}):
\begin{lstlisting}
	# Import the class vector from python
	from vpython import vector
	# Initialize distance and velocity vectors
	vec_rM0 = vector(0, rM0, 0)
	vec_vM0 = vector(vm0, 0, 0)
\end{lstlisting}
According to Eq.~(\ref{eq:newton}), the force which acts on mercury changes the velocity of mercury which eventually changes the position. However, we first need to fix the time step $\Delta t$ and use the estimate
of Eq.~(\ref{eq:check}) as a guidance {\bf Careful: Check units of dt}
\begin{lstlisting}
	# Definition of the time step
        dt = 2 * vm0 / aM / 100
\end{lstlisting}
Here the factor $1/100$ makes sure that $\Delta t$ is indeed
consistent with Eq.~(\ref{eq:check}).  {\bf We need to make sure that 'dt' is known
to the subroutines}
Now we are in the position to calculate
location and velocity of the planet at $t_0+\Delta t$ using the following comands
\begin{lstlisting}
	# Compute the strength of the acceleration
	aMS = aM * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
	# Multiply by the direction
	vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
	# Update velocity vector
	vec_vM_new = vec_vM_old + vec_aMS * dt
	# Update position vector
	vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note the beauty of working with the predefined \texttt{vector} class: the basic vector operations are already implemented.  The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.  Also the magnitude of a vector -- \texttt{vector.mag} -- is a property of the vector and can be easily extracted.

@@

Suppose we want to save the number of turns in the variable $counter$, the maximal number of turns in $maxcounter$ (here e.g. 10), the initial aphelion in the vector $a0$ and the final aphelion in the vector $a$. In addition to find out when the aphelion is reached, the last two positions $rlast$ and $rbeforelast$ have to be known. To set this up, before the while loop the following code needs to be included
\begin{lstlisting}
	r=r0
	rlast=r0
	counter=-1
	maxcounter=10
\end{lstlisting}
Then the following code needs to be included in the $while$-loop
\begin{lstlisting}
    rbeforelast=rlast
    rlast=r
    r=r+v*dt
    if (counter==-1): a0=rlast
    if (rbeforelast.mag<rlast.mag and rlast.mag>r.mag): counter=counter+1
    if counter==maxcounter:
    	a=rlast
    	break
\end{lstlisting}
and the condition for the $while$-loop should be set to $True$.
The angle between two vectors is given by the formula
 \begin{equation}
 	\sphericalangle(\vec{a}_0,\vec{a}) = \cos^{-1} \left( \frac{\vec{a}_0 \cdot \vec{a}}{|\vec{a}_0|\:|\vec{a}|} \right)
 \end{equation}
This is readily implemented in \texttt{Vpython} via
\begin{lstlisting}
	angle=acos(dot(a0,a)/(a0.mag*a.mag))/maxcounter*180/pi
	print("angle: ", (angle))
\end{lstlisting}

@@

Instead of providing the template to the students, they could build the program from scratch. Of course, this requires some basic knowledge in \texttt{VPython}, as they could acquire for example by working through a ball in the box example or the like. They could even look up the needed parameters on their own.
\item \textbf{Why can we work in one plane?}

@@

\appendix
\section{Instructions to install VPython}
\label{appendix:python}
